---
title: "Prosper Loan Data"
author: "Noah Zheng"
date: "January 9, 2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: lumen
    highlight: textmate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=10,fig.height=7,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

```



# Prosper Loan Project 

## Background Information 
The dataset is from Prosper, an American company in the peer-to-peer lending industry. Prosper maintains a full public database of all loans issued through its marketplace on its website. This database and all market statistics can be accessed and queried for analysis of loan performance over time. 

## Source 
1. https://en.wikipedia.org/wiki/Prosper_Marketplace#cite_note-Prosper_Marketplace,_Inc-6 
2. https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0 
3. https://www.wikihow.com/Invest-on-Prosper.com

## Process
1. Overview of the data
2. Analysis
  * Univariateanalysis and plots
  * Bivariate analysis and plots
  * Multivariate analysis and plots
3. Final plots and summary
4. reflection


## Overview of the Data

From Udacity:

*This data set contains 113,937 loans with 81 variables on each loan, including loan amount, borrower rate (or interest rate), current loan status, borrower income, borrower employment status, borrower credit history, and the latest payment information.*

### Main Features of Interest

- ListingCreationDate
- Term
- LoanStatus
- BorrowerRate
- LeanderYield
- ProsperRating(numeric)
- ProsperScore
- ListingCategory
- BorrowerState
- IsBorrowerHomeowner
- DebtToIncomeRatio
- LoanOriginalAmount 
- Investors


## Exploratory Analysis

```{r }
#import the data
dd <- read.csv("prosperLoanData.csv")

```

### Univariate analysis 

```{r }
#Select columns
library(dplyr)
cols <- c('ListingCreationDate', 'Term', 'LoanStatus', 'BorrowerRate',
          'LenderYield', 'ProsperRating..numeric.', 'ProsperScore',
          'ListingCategory..numeric.', 'BorrowerState',
          'IsBorrowerHomeowner', 'DebtToIncomeRatio',
          'LoanOriginalAmount', 'Investors')

dd_s <- select(dd, cols)

```
#### The Structure of the Data 

```{r }
#structure of the data
str(dd_s)
summary(dd_s)
```

Firtly, I am interested in the listing over the time. Therefore, I generate the following plot.

```{r }
#ggpair, univariate plot
library(ggplot2)

#convert date time to date
library(anytime)
dd_s$ListingCreationDate <- anytime(dd_s$ListingCreationDate)
dd_s$ListingCreationDate <- as.Date(dd_s$ListingCreationDate, format = "%Y/%m")

#str(dd_s)

#https://stackoverflow.com/questions/11748384/formatting-dates-on-x-axis-in-ggplot2
#https://stackoverflow.com/questions/41855673/r-ggplot-display-all-dates-on-x-axis
library(scales)
library(ggpubr)
##1
ggplot(dd_s, aes(x = ListingCreationDate)) +
  geom_histogram() +
  theme_bw() +
  labs(x = 'month', y = 'Number of Listing') +
  scale_x_date(labels = date_format("%y-%m"), date_breaks = "2 month") +
  ggpubr::rotate_x_text()

```

It can be noticed that there is a big drop in listing from 2008-09. It takes almost 3 years for the listing to recover to 2008-07 level. What happened in September of 2008 relating to finance is the collapse of Lehman Brothers on September 15. Another information is of interest is that the Prosper rating is introduced in July 2009, when the listing starts to increase. Another sharp decline happened in 2014-03. It can be explored further with other information. 


#### The Term of Loans 

```{r}
##2 bar plot of terms
ggplot(dd_s, aes(x=factor(Term))) +
  geom_bar()
```
 
The majority of loans are with a term of 36 months. Loans with 12 months are quite small. Around 25000 loans are with a 60 months term. 


#### Loan Status
```{r}
##3 bar plot of loan status
ggplot(dd_s, aes(x = LoanStatus)) +
  geom_bar() + 
  geom_text(stat = 'count', aes(label = ..count..), hjust=-0.15) +
  coord_flip(ylim = c(0,62000))
```

The majority of loans are current. The second is completed. The third is charged off. The fourth is defaulted. 


#### The Borrower Rate

```{r}
##4 histogram of borrower rate
ggplot(dd_s, aes(x = BorrowerRate)) +
  geom_histogram()
```

The distribution of borrower rate is right skewed with median 0.1840 < mean 0.1928. 

```{r}
summary(dd_s$BorrowerRate)
```


#### Lender Yield
```{r}
##5 histogram of lender yield
ggplot(dd_s, aes(x = LenderYield)) +
  geom_histogram()
```

As lender yield is equal to the interest rate on the loan less the servicing fee, the shape is essentially the same as borrower rate. 

```{r}
summary(dd_s$LenderYield)
```

#### Prosper Rating 

```{r}
##6 bar plot of prosper rating
ggplot(dd_s, aes(x=factor(ProsperRating..numeric.))) +
  geom_bar()
```
 
The distribution of rating is close to a normal distribution except for the NA which is because that the Prosper rating is introduced in July 2009.

#### Prosper Score
```{r}
##7 histogram of prosper score
ggplot(dd_s, aes(x = ProsperScore)) +
  geom_histogram() +
  scale_x_continuous(breaks = seq(1,11,1))
```

The distributinon of Prosper Score is of bell shape with a median of 6 and mean of 5.95.

```{r}
summary(dd_s$ProsperScore)
```

#### Listing Category

The category of the listing that the borrower selected when posting their listing: 0 - Not Available, 1 - Debt Consolidation, 2 - Home Improvement, 3 - Business, 4 - Personal Loan, 5 - Student Use, 6 - Auto, 7- Other, 8 - Baby&Adoption, 9 - Boat, 10 - Cosmetic Procedure, 11 - Engagement Ring, 12 - Green Loans, 13 - Household Expenses, 14 - Large Purchases, 15 - Medical/Dental, 16 - Motorcycle, 17 - RV, 18 - Taxes, 19 - Vacation, 20 - Wedding Loans   

```{r}
#https://stackoverflow.com/questions/10002536/how-do-i-replace-numeric-codes-with-value-labels-from-a-lookup-table
# convert listing categories into readable names
a <- 0:20
names(a) <- c('Not Available', 'Debt Consolidation', 'Home Improvement', 'Business', 'Personal Loan', 'Student Use', 'Auto', 'Other', 'Baby&Adoption', 'Boat', 'Cosmetic Procedure', 'Engagement Ring', 'Green Loans', 'Household Expenses', 'Large Purchases', 'Medical/Dental', 'Motorcycle', 'RV', 'Taxes', 'Vacation', 'Wedding Loans')

dd_s$ListingCategory..numeric. <- names(a)[match(dd_s$ListingCategory..numeric., a)]
```


```{r}
#https://stackoverflow.com/questions/37853539/how-to-draw-sorted-frequency-barplot-with-ggplot-in-r
##8 bar plot of listing category
ggplot(dd_s, aes(x = reorder(ListingCategory..numeric., ListingCategory..numeric., function(x) -length(x)))) +
  geom_bar() +
  xlab("Category") +
  ggpubr::rotate_x_text()

```

Most of the listing falls in Debt Consolidation. The second category that is useful is Home Improvement. The third is Business. 


#### Borrowing State

```{r}
##9 bar plot of borrwoing state
ggplot(dd_s, 
       aes(x=reorder(BorrowerState, BorrowerState, function(x) - length(x))
           )
       ) +
  xlab('state') +
  geom_bar()
```

California has the most listings. The second is Texas. The third is New York.

#### Borrower a Home Owner

```{r}
##10 bar plot of home owner
ggplot(dd_s, aes(x = IsBorrowerHomeowner)) +
  geom_bar()

```

There are around 1000 more home owners.

```{r}
summary(dd_s$IsBorrowerHomeowner)
```


#### Debt to Income Ratio

```{r}
##11 histogram of debt to income ratio
ggplot(dd_s, aes(x = DebtToIncomeRatio)) +
  geom_histogram()
```

More than 75% of the listings have a debt to income ratio less than 0.32.

```{r}
summary(dd_s$DebtToIncomeRatio)
```

#### Loan Amount 

```{r }
##12 plot of LoanOriginalAmount
ggplot(dd_s, aes(x = LoanOriginalAmount)) +
  geom_histogram()
```
  
It shows a right skewed trend, mean 8447 > median 6500. 

```{r}
summary(dd_s$LoanOriginalAmount)
```

#### Investors
```{r}
##13 histogram of investors 
ggplot(dd_s, aes(x = Investors)) +
  geom_histogram()
```

It is right skewed with a median of 44 and mean of 80.48.

```{r}
summary(dd_s$Investors)
```

### Univariate Analysis

#### The Relationship between the Yield and Rating 

```{r }
##14
ggplot(na.omit(filter(dd_s)), aes(as.factor(ProsperRating..numeric.), LenderYield)) +
  geom_boxplot(alpha=0.4, color="orange") +
  geom_point(stat = "summary", fun.y = mean, size = 4, shape=23)
```
 
It can be noticed that the yield decreases when rating increases.


What is the trend when grouped by different loan status?

```{r }
##15
ggplot(dd_s, aes(ProsperRating..numeric., LenderYield)) +
  geom_line(aes(color = LoanStatus), stat = "summary", fun.y = mean)
```

They all share the same trend, lender yield decreasing when prosper rating increasing. 


#### Exploring the Relationship between Yield and Loan Amount 


```{r }
##16  
ggplot(dd_s, aes(LoanOriginalAmount, LenderYield)) +
  geom_point(alpha = 1/50, position = position_jitter())+
  geom_smooth(color = "red")
```

No clear pattern. It seems that there is no clear relationship between loan amount and lender yield. I supposed that the lender yield may increase when the loan amount increases. 


#### The Relationship between Loan Amount and Rating. 

```{r }
##17
dd_s <- dd_s %>% mutate(
  Status = ifelse(
    LoanStatus %in% c('Chargedoff', 'Defaulted'), 0,
    ifelse(
      LoanStatus %in% 
        c("Completed", "Current", "FinalPaymentInProgress"), 1,
      ifelse(
        LoanStatus %in% 'Canclled', 2, 3
      )
    )
  )
)

dd_s$Status <- factor(dd_s$Status, levels = 0:3, 
                         labels = c("Defaulted", 
                                    "Current or paid", 
                                    "Canclled",
                                    "Past due"))


ggplot(na.omit(filter(dd_s)), aes(ProsperRating..numeric., LoanOriginalAmount)) +
  geom_point(alpha = 1/100, position = position_jitter()) +
  geom_line(aes(color = Status), stat = "summary", fun.y = "mean") +
  ylim(0,quantile(dd_s$LoanOriginalAmount, .99))
```

Listings with high risk tends to have low amount than that with low risk. It is probably that the borrowers with high risk know they cannot get a big loan. Defaulted loan has less loan amount on average than current or paid and past due loans. 


#### Listing Category and Loan Amount

```{r}
##18 
ggplot(na.omit(filter(dd_s)), aes(x = ListingCategory..numeric., y = LoanOriginalAmount)) +
  geom_boxplot() +
  ggpubr::rotate_x_text()
```

The top two categories that have the highest median loan amount are debt consolidation and baby and adoption. The third is wedding loans. It is interesting to find out that baby and adoption and wedding are expensive.

#### Loan Amount and Investors

```{r}
##19
ggplot(na.omit(filter(dd_s)), aes(x = LoanOriginalAmount, y = Investors)) +
  geom_point() +
  geom_smooth(color = 'orange')
```

The line indicates the trend of number of investors along the loan amount. After 20000, the trend of the number of investors remains stable. 
### Multivariate Plots

#### Lender Yield, Prosper Rate and Term
```{r}
##20 
ggplot(na.omit(filter(dd_s, DebtToIncomeRatio < 1)), 
               aes(ProsperRating..numeric., LenderYield,                                   color=ProsperRating..numeric.)) +
    geom_point(alpha = 0.8, position = "jitter") +
    facet_grid(.~ Term)
```

It can be noticed that the lender yield all exhibit the same trend in three term categories. Lender yield decreases when Prosper rating increases. 

## Final Plots and Summary

### Plot 1
```{r }
ggplot(dd_s, aes(x = ListingCreationDate)) +
  geom_histogram() +
  theme_bw() +
  labs(x = 'month', y = 'Number of Listing') +
  scale_x_date(labels = date_format("%y-%m"), date_breaks = "2 month") +
  ggtitle("Number of Listing Over the Time") +
  ggpubr::rotate_x_text()

```

This is the plot of number of listing spread over the period from July 2005 to Sept. 2014. There is a deep decrease in the Q4 2018. What happened then is the financial crisis.The listing started to recover around July 2009 when Prosper introduced the Prosper rating. The listing started to prosper around Jan. 2013 and reached its highest in Nov. 2013 and then declined. 

### Plot 2
```{r }
ggplot(dd_s, aes(ProsperRating..numeric., LenderYield)) +
  geom_line(aes(color = LoanStatus), stat = "summary", fun.y = mean)+
  ggtitle("The Relationship between Lender Yield and Prosper Rating") +
  ylab("Lender Yield")
```

This is the plot of the relationship between lender yield and Propser rating. It indicates that, no matter what loan status, higher Prosper rating means lower lender yield. The average of lender yield is above 0.3 with rating of 1 and it decreases to around 0.1 with rating of 6.

### Plot 3
```{r }
ggplot(na.omit(filter(dd_s)), aes(x = LoanOriginalAmount, y = Investors)) +
  geom_point() +
  geom_smooth(color = 'orange') +
  ggtitle("Relationship between Number of Investor and Loan Amount") +
  ylab("Number of Investers")
```

THis is the plot of the relationshop between number of investors and loan amount. I supposed that the larger loan amount, the more investors. The plot indicates that there is quite weak relationship between these two factors. 

## Reflection

I spent a lot of time exploring the dataset to understanding the meaning of each variables. This dataset has 81 variables and I have to understanding all of them and then to choose the variables of interest for data analysis.


I got lost when exploring some of the variables at times. There are no obvious trend in some of the variable and I have to go back and forth to try the coorrelation between different variables. 

As shown in Plot 1, there is a obvious decline in Q4 2008 and Prosper has made some change in July 2009. There should be some other information that can reveal more about how Prosper made the change. I tried but failed. I realised that if you want to dig into the data to find more useful insight you have to understanding data. Otherwise, R can only help to find some obvious observation such as the positive relationship between the Prosper rating and lender yield. 



 

