---
title: "Loan_Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loan Project 

## Background Information 

## Source 
1. https://en.wikipedia.org/wiki/Prosper_Marketplace#cite_note-Prosper_Marketplace,_Inc-6 
2. https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0 
3. https://www.wikihow.com/Invest-on-Prosper.com

## Question
It is so hard to ask the right question. Assume I was a lender. What I care about is the risk of a list and the yield. When I was browsing the lists, what factor should I care most? Is expected return reliable? The income of the borrower information is not reliable as borrower would likely to appreciate their income in order to get a loan. Therefore, reliable factors, such as state, credit profile, term, prosperscore, inquiriesLast6Months/TotalInquiries (too much inquiries might indicate high risk), currentDelinquencies, totalProsperLoans,   should be choosen. 
1. Which state are more liable?
2. 
3. Which month has the highest demand of loan (out of curiosity)
4. The relationship between Loan status and ProsperRating
5. Predict LenderYield 
6. Can I just rely on teh credit profile to lend money? (analyse the relatinship between credit score and loan status)
7. High rish high return? (low credit score, loan status bad?)
8. 


```{r}
dd <- read.csv("prosperLoanData.csv")
str(dd)
head(dd)
summary(dd$ListingCreationDate)
str(dd$ListingCreationDate)
summary(dd$LoanOriginationDate)
range(dd$LoanOriginationDate)
```

```{r}
#Select data, trim the date with only date
```

